---
format:
  dashboard:
    theme: minty
    orientation: rows
    scrolling: true
    logo: logo/plastic-waste-icon-vector.jpg
    nav-buttons:
      - icon: envelope
        href: mailto:mpsouza@calpoly.edu 
      - icon: linkedin
        href: https://www.linkedin.com/in/mason-souza-19082a293/
---

# Research Question 2

```{r}
#| include: false

library(tidyverse)
library(httr)
library(jsonlite)
library(tidyjson)
library(glue)
library(countrycode)
library(gt)
library(hrbrthemes)
library(viridis)
library(hexbin)

# Load plastic waste data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

# Add ISO-2 codes
plastics <- plastics %>%
  filter(country != "EMPTY") %>%
  mutate(iso2 = countrycode(country, origin = "country.name", destination = "iso2c"))

iso2_list <- unique(plastics$iso2)
iso2_list <- iso2_list[iso2_list != "TW"]

# GDP API function
get_gdp <- function(iso2) {
  url <- glue("https://api.worldbank.org/v2/country/{iso2}/indicator/NY.GDP.MKTP.CD?format=json")
  my_data <- fromJSON(url)
  if (is.null(my_data[[2]])) {
    return(NULL)
  } else {
    return(my_data[[2]] %>%
             filter(date %in% c("2019", "2020")) %>%
             select(date, value) %>%
             mutate(iso2c = iso2, value = value / 1e9) %>%
             rename(gdp_billions = value))
  }
}

gdp_data <- map_dfr(iso2_list, get_gdp)

# Population API function
get_population <- function(iso2) {
  url <- glue("https://api.worldbank.org/v2/country/{iso2}/indicator/SP.POP.TOTL?format=json")
  my_data <- fromJSON(url)
  if (is.null(my_data[[2]])) {
    return(NULL)
  } else {
    return(my_data[[2]] %>%
             filter(date %in% c("2019", "2020")) %>%
             select(date, value) %>%
             mutate(iso2c = iso2))
  }
}

population_data <- map_dfr(iso2_list, get_population)

# Load CO2 data
co2 <- readr::read_csv(
  "https://raw.githubusercontent.com/owid/co2-data/master/owid-co2-data.csv")

# Build combined meta dataset
co2_sm <- co2 %>%
  filter(year %in% c(2019, 2020)) %>%
  mutate(iso2 = countrycode(iso_code, "iso3c", "iso2c"))

gdp_data <- gdp_data %>% mutate(date = as.numeric(date))
population_data <- population_data %>% mutate(date = as.numeric(date))

meta_set <- plastics %>%
  left_join(gdp_data, by = c("iso2" = "iso2c", "year" = "date")) %>%
  left_join(population_data, by = c("iso2" = "iso2c", "year" = "date")) %>%
  left_join(co2_sm, by = c("iso2", "year")) %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  )

meta_set <- meta_set %>%
  mutate(
    parent_company = ifelse(
      parent_company %in% c("NULL", "null"),
      "Grand Total",
      parent_company
    )
  )

model_data <- meta_set %>%
  filter(
    year == 2019,
    parent_company == "Grand Total"
  ) %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  ) %>%
  filter(
    !is.na(gdp_per_capita),
    !is.na(plastic_waste_per_capita),
    gdp_per_capita > 0,
    plastic_waste_per_capita > 0
  )
```

## Data Sources

### 🗑️ Plastic Waste Data

***🗑️ Plastic Waste Data***
This data set was independently gathered by Sarah Sauve to contribute to the global audit that the Break Free from Plastic organization holds each year. Data was collected in St. John, Newfoundland from 2019 to 2020. The data set includes counts of different plastic types (polyester, PVC, polyethylene, etc.), the grand total of all plastic types, number of counting events, and number of volunteers at each cleanup.

### 🌍 Population & GDP Data
***🌍 Population & GDP Data***
Through the World Bank Group's API, we collected population and GDP data for each country in our plastic waste data set for 2019 and 2020. The World Bank coordinates and compiles data from National Statistics Agencies, Central Banks, and Government Agencies into accessible APIs. Data are gathered via household surveys, digital records, censuses, and other secondary sources.

### 🏭 CO₂ Emissions Data
***🏭 CO₂ Emissions Data***
Our final data set contains country-level CO₂ emissions published by *Our World in Data*, sourced from the Global Carbon Project and other scientific databases, spanning 1750–2024. We restricted it to 2019–2020 to match our other data. CO₂ estimates are derived by tracking a country's net fuel consumption (extracted + imported − exported) and multiplying each fuel type by its emission factor to compute total emissions.

## Research Questions

### 🔬 Our Research Questions

::: {style="font-size: 1.05em; line-height: 2;"}
**Question 1 — GDP & Plastic Waste**

What is the relationship between GDP per capita and plastic waste generation across countries in 2019–2020? Did the 2020 pandemic have the same effect on both metrics?

---

**Question 2 — GDP & CO₂ Emissions**

What is the relationship between GDP per capita and CO₂ emissions per capita across countries in 2019? How does population size influence this relationship?
:::



## Data Summaries {.tabset}

### 🌍 GDP & CO₂ by Population

Is there a visible trend between countries' per capita GDP, CO₂ emissions, and population size?

```{r}
country_gdp_co2_summary <- function(data) {
  data |>
    filter(
      year == 2019,
      !is.na(country.x),
      country.x != "Taiwan_ Republic of China (ROC)"
    ) |>
    group_by(country.x) |>
    summarise(
      gdp_per_capita = mean(gdp_billions * 1e9 / population, na.rm = TRUE),
      co2_per_capita = mean(co2_per_capita, na.rm = TRUE),
      population = mean(population, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(population))
}

gdp_co2_table <- country_gdp_co2_summary(meta_set)

gdp_co2_table |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    country.x = "Country",
    gdp_per_capita = "GDP Per Capita",
    co2_per_capita = "CO₂ per Capita",
    population = "Population"
  ) |>
  fmt_currency(
    columns = contains("gdp"),
    currency = "USD",
    decimals = 0
  ) |>
  fmt_number(
    columns = c(co2_per_capita, population),
    decimals = 0,
    use_seps = TRUE
  ) |>
  tab_header(
    title = md("**GDP vs CO₂ Emissions per Capita (2019)**"),
    subtitle = "Top 10 countries sorted by population"
  ) |>
  tab_caption(
    "Countries GDP per capita & CO₂ per capita, sorted by population"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

### 📈 GDP & CO₂ Rankings

Does there appear to be a relationship between the rank of a country's GDP per capita and their CO₂ per capita?

```{r}
countries <- meta_set |>
  filter(
    year == 2019,
    !is.na(country.x),
    country.x != "Taiwan"
  ) |>
  distinct(country.x) |>
  pull(country.x)

ranked_table <- map_dfr(countries, function(c) {
  meta_set |>
    filter(country.x == c, year == 2019) |>
    summarise(
      country = c,
      gdp_per_capita = mean(gdp_billions * 1e9 / population, na.rm = TRUE),
      co2_per_capita = mean(co2_per_capita, na.rm = TRUE)
    )
}) |>
  filter(!is.na(gdp_per_capita), !is.na(co2_per_capita)) |>
  mutate(
    gdp_rank = min_rank(desc(gdp_per_capita)),
    co2_rank = min_rank(desc(co2_per_capita))
  ) |>
  arrange(gdp_rank)

ranked_table |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    country = "Country",
    gdp_per_capita = "GDP Per Capita",
    co2_per_capita = "CO₂ Per Capita",
    gdp_rank = "GDP Rank",
    co2_rank = "CO₂ Rank"
  ) |>
  fmt_currency(
    columns = contains("gdp"),
    currency = "USD",
    decimals = 0
  ) |>
  fmt_number(
    columns = c(co2_per_capita),
    decimals = 2,
    use_seps = TRUE
  ) |>
  fmt_number(
    columns = c(gdp_rank, co2_rank),
    decimals = 0
  ) |>
  tab_spanner(
    label = "Per-Capita Measures",
    columns = c(gdp_per_capita, co2_per_capita)
  ) |>
  tab_spanner(
    label = "Global Rankings",
    columns = c(gdp_rank, co2_rank)
  ) |>
  data_color(
    columns = c(gdp_rank, co2_rank),
    fn = scales::col_numeric(
      palette = c("#DCEAF7", "#2C7FB8"),
      domain = c(1, 30)
    )
  ) |>
  tab_header(
    title = md("**GDP and CO₂ Per-Capita Ranking Comparison**"),
    subtitle = "Top 10 countries ranked by GDP per capita in 2019"
  ) |>
  tab_caption(
    "Table 6. Comparison of country rankings for GDP per capita and CO₂ emissions per capita in 2019."
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## Analysis

CO₂ and GDP rank appear to move together; countries with the highest GDP per capita tend to also rank highly in CO₂ emissions. This aligns with the positive relationship found in the regression model.

## 🫧 Bubble Plot: GDP vs CO₂ Emissions Per Capita

```{r}
#| warning: false
#| label: fig-rq2-bubble
#| fig-cap: "GDP per capita and CO2 emissions per capita across countries in 2019, with bubble size representing population."
#| fig-alt: "Bubble plot of GDP per capita versus CO2 emissions per capita for 51 countries in 2019. Bubble size represents population and color indicates GDP level. Countries with higher GDP per capita tend to also have higher CO2 emissions per capita."

model_data %>%
  arrange(desc(population)) %>%
  mutate(
    country.x = as.factor(country.x),
    gdp_level = case_when(
      is.na(gdp_billions) ~ "Other",
      gdp_billions < 100 ~ "Low",
      gdp_billions < 500 ~ "Lower-Mid",
      gdp_billions < 1000 ~ "Upper-Mid",
      gdp_billions >= 1000 ~ "High"
    ),
    gdp_level = factor(
      gdp_level,
      levels = c("Low", "Lower-Mid", "Upper-Mid", "High", "Other")
    )
  ) %>%
  ggplot(aes(
    x = gdp_per_capita,
    y = co2_per_capita,
    size = population,
    color = gdp_level
  )) +
  geom_point(alpha = 0.65) +
  scale_size(range = c(2, 14), name = "Population") +
  scale_color_viridis_d(option = "A", end = 0.9, name = "GDP Level") +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_log10(labels = scales::label_number()) +
  theme_ft_rc() +
  labs(
    x = "GDP Per Capita",
    y = "CO2 Emissions Per Capita",
    title = "Bubble Plot of GDP and CO2 Emissions Per Capita",
    subtitle = "Countries in 2019, with bubble size representing population"
  )
```

## 📊 GDP vs Plastic Waste Per Capita (2019 & 2020)

```{r}
#| warning: false
#| label: fig-rq1-facet
#| fig-cap: "Relationship between GDP per capita and CO2 waste per capita across countries in 2019 and 2020, with separate panels for each year."
#| fig-alt: "Faceted scatterplot showing GDP per capita versus CO2 waste per capita for countries in 2019 and 2020. Each panel includes a fitted trend line. The relationship appears positive in both years, allowing comparison of how the pattern changes between 2019 and 2020."

ggplot(
  meta_set %>%
    filter(year %in% c(2019, 2020), parent_company == "Grand Total"),
  aes(x = gdp_per_capita, y = co2_per_capita)
) +
  geom_point(color = "#2C7FB8", alpha = 0.7, size = 2.5) +
  geom_smooth(method = "loess", se = FALSE, color = "#D95F0E", linewidth = 1) +
  scale_x_log10(labels = scales::label_dollar()) +
  scale_y_log10(labels = scales::label_number()) +
  facet_wrap(~year) +
  labs(
    title = "GDP per Capita vs CO2 Waste per Capita",
    subtitle = "Comparison of country-level relationships in 2019 and 2020",
    x = "GDP per Capita",
    y = "CO2 Waste per Capita"
  ) +
  theme_ft_rc()
```

#Mystery Country???

### We need a country that is one of: An outlier, A representative case, or A case of change (need to know we can find country-specific data suitable for a dynamic visualization). We also need to include a short “Why this country?” section (2–3 sentences) on this dashboard page. Also need a dynamic visualization and a leaflet map.