---
title: "Break Free From Plastic Analysis"
format:
  dashboard:
    theme: minty
    orientation: rows
    fit: true
    scrolling: true
    logo: logo/plastic-waste-icon-vector.jpg
    nav-buttons:
      - icon: envelope
        href: mailto:mpsouza@calpoly.edu 
      - icon: linkedin
        href: https://www.linkedin.com/in/mason-souza-19082a293/
server: shiny 
---
```{r}
#| include: false
#| context: setup
#| label: load-packages

library(tidyverse)
library(httr)
library(jsonlite)
library(tidyjson)
library(glue)
library(countrycode)
library(gt)
library(hrbrthemes)
library(viridis)
library(hexbin)
library(leaflet)
library(plotly)
```

```{r}
#| context: data

# Load plastic waste data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')

# Add ISO-2 codes
plastics <- plastics %>%
  filter(country != "EMPTY") %>%
  mutate(iso2 = countrycode(country, origin = "country.name", destination = "iso2c"))

iso2_list <- unique(plastics$iso2)
iso2_list <- iso2_list[iso2_list != "TW"]

# GDP API function
get_gdp <- function(iso2) {
  url <- glue("https://api.worldbank.org/v2/country/{iso2}/indicator/NY.GDP.MKTP.CD?format=json")
  my_data <- fromJSON(url)
  if (is.null(my_data[[2]])) {
    return(NULL)
  } else {
    return(my_data[[2]] %>%
             filter(date %in% c("2019", "2020")) %>%
             select(date, value) %>%
             mutate(iso2c = iso2, value = value / 1e9) %>%
             rename(gdp_billions = value))
  }
}

gdp_data <- map_dfr(iso2_list, get_gdp)
Sys.sleep(1)

# Population API function
get_population <- function(iso2) {
  url <- glue("https://api.worldbank.org/v2/country/{iso2}/indicator/SP.POP.TOTL?format=json")
  my_data <- fromJSON(url)
  if (is.null(my_data[[2]])) {
    return(NULL)
  } else {
    return(my_data[[2]] %>%
             filter(date %in% c("2019", "2020")) %>%
             select(date, value) %>%
             mutate(iso2c = iso2))
  }
}

population_data <- map_dfr(iso2_list, get_population)

# Load CO2 data
co2 <- readr::read_csv(
  "https://raw.githubusercontent.com/owid/co2-data/master/owid-co2-data.csv")

# Build combined meta dataset
co2_sm <- co2 %>%
  filter(year %in% c(2019, 2020)) %>%
  mutate(iso2 = countrycode(iso_code, "iso3c", "iso2c"))

gdp_data <- gdp_data %>% mutate(date = as.numeric(date))
population_data <- population_data %>% mutate(date = as.numeric(date))

meta_set <- plastics %>%
  left_join(gdp_data, by = c("iso2" = "iso2c", "year" = "date")) %>%
  left_join(population_data, by = c("iso2" = "iso2c", "year" = "date")) %>%
  left_join(co2_sm, by = c("iso2", "year")) %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  )

meta_set <- meta_set %>%
  mutate(
    parent_company = ifelse(
      parent_company %in% c("NULL", "null"),
      "Grand Total",
      parent_company
    )
  )

model_data <- meta_set %>%
  filter(
    year == 2019,
    parent_company == "Grand Total"
  ) %>%
  mutate(
    gdp_per_capita = gdp_billions * 1e9 / population,
    plastic_waste_per_capita = grand_total / population
  ) %>%
  filter(
    !is.na(gdp_per_capita),
    !is.na(plastic_waste_per_capita),
    gdp_per_capita > 0,
    plastic_waste_per_capita > 0
  )


#New Luxembourg data
luxembourg_data <- readLines("commuters-to-luxembourg-2019-2018-commuters-lux-2019-0.geojson")

# Gapminder-style animated dataset (1970–2022)
gapminder_co2 <- co2 |>
  filter(
    year >= 1970,
    !is.na(iso_code),
    iso_code != "",
    !str_starts(iso_code, "OWID")
  ) |>
  mutate(
    gdp_per_capita = gdp / population,
    continent = countrycode(iso_code, "iso3c", "continent")
  ) |>
  filter(
    !is.na(gdp_per_capita),
    !is.na(co2_per_capita),
    !is.na(population),
    !is.na(continent),
    gdp_per_capita > 0,
    co2_per_capita > 0
  ) |>
  select(country, year, gdp_per_capita, co2_per_capita, population, continent)
```

# Analyzing GDP and CO₂ Emissions

## Row {height="150px"}

### 🌍 Population & GDP Dataa

***🌍 Population & GDP Data***
Through the World Bank Group's API, we collected population and GDP data for each country in our plastic waste data set for 2019 and 2020. The World Bank coordinates and compiles data from National Statistics Agencies, Central Banks, and Government Agencies into accessible APIs. Data are gathered via household surveys, digital records, censuses, and other secondary sources.

### 🏭 CO₂ Emissions Data

***🏭 CO₂ Emissions Data***
Our final data set contains country-level CO₂ emissions published by *Our World in Data*, sourced from the Global Carbon Project and other scientific databases, spanning 1750–2024. We restricted it to 2019–2020 to match our other data. CO₂ estimates are derived by tracking a country's net fuel consumption (extracted + imported − exported) and multiplying each fuel type by its emission factor to compute total emissions.

## Row {height="600px"}

### 📋 Data Summaries {.tabset}

#### 🌍 GDP & CO₂ by Population


```{r}
#| title: Is there a visible trend between countries' per capita GDP, CO₂ emissions, and population size?
country_gdp_co2_summary <- function(data) {
  data |>
    filter(
      year == 2019,
      !is.na(country.x),
      country.x != "Taiwan_ Republic of China (ROC)"
    ) |>
    group_by(country.x) |>
    summarise(
      gdp_per_capita = mean(gdp_billions * 1e9 / population, na.rm = TRUE),
      co2_per_capita = mean(co2_per_capita, na.rm = TRUE),
      population = mean(population, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(population))
}

gdp_co2_table <- country_gdp_co2_summary(meta_set)

gdp_co2_table |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    country.x = "Country",
    gdp_per_capita = "GDP Per Capita",
    co2_per_capita = "CO₂ per Capita",
    population = "Population"
  ) |>
  fmt_currency(
    columns = contains("gdp"),
    currency = "USD",
    decimals = 0
  ) |>
  fmt_number(
    columns = c(co2_per_capita, population),
    decimals = 0,
    use_seps = TRUE
  ) |>
  tab_header(
    title = md("**GDP vs CO₂ Emissions per Capita (2019)**"),
    subtitle = "Top 10 countries sorted by population"
  ) |>
  tab_caption(
    "Countries GDP per capita & CO₂ per capita, sorted by population"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

#### 📈 GDP & CO₂ Rankings


```{r}
#| title: Does there appear to be a relationship between the rank of a country's GDP per capita and their CO₂ per capita?


countries <- meta_set |>
  filter(
    year == 2019,
    !is.na(country.x),
    country.x != "Taiwan"
  ) |>
  distinct(country.x) |>
  pull(country.x)

ranked_table <- map_dfr(countries, function(c) {
  meta_set |>
    filter(country.x == c, year == 2019) |>
    summarise(
      country = c,
      gdp_per_capita = mean(gdp_billions * 1e9 / population, na.rm = TRUE),
      co2_per_capita = mean(co2_per_capita, na.rm = TRUE)
    )
}) |>
  filter(!is.na(gdp_per_capita), !is.na(co2_per_capita)) |>
  mutate(
    gdp_rank = min_rank(desc(gdp_per_capita)),
    co2_rank = min_rank(desc(co2_per_capita))
  ) |>
  arrange(gdp_rank)

ranked_table |>
  slice_head(n = 10) |>
  gt() |>
  cols_label(
    country = "Country",
    gdp_per_capita = "GDP Per Capita",
    co2_per_capita = "CO₂ Per Capita",
    gdp_rank = "GDP Rank",
    co2_rank = "CO₂ Rank"
  ) |>
  fmt_currency(
    columns = contains("gdp"),
    currency = "USD",
    decimals = 0
  ) |>
  fmt_number(
    columns = c(co2_per_capita),
    decimals = 2,
    use_seps = TRUE
  ) |>
  fmt_number(
    columns = c(gdp_rank, co2_rank),
    decimals = 0
  ) |>
  tab_spanner(
    label = "Per-Capita Measures",
    columns = c(gdp_per_capita, co2_per_capita)
  ) |>
  tab_spanner(
    label = "Global Rankings",
    columns = c(gdp_rank, co2_rank)
  ) |>
  data_color(
    columns = c(gdp_rank, co2_rank),
    fn = scales::col_numeric(
      palette = c("#DCEAF7", "#2C7FB8"),
      domain = c(1, 30)
    )
  ) |>
  tab_header(
    title = md("**GDP and CO₂ Per-Capita Ranking Comparison**"),
    subtitle = "Top 10 countries ranked by GDP per capita in 2019"
  ) |>
  tab_caption(
    "Table 6. Comparison of country rankings for GDP per capita and CO₂ emissions per capita in 2019."
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```



### 📊 Visualizations {.tabset}

#### 🫧 Interactive Bubble Plot

```{r}
#| content: card-sidebar

selectInput("x_variables", "Select X Variable", choices = c(
    "GDP per Capita" = "gdp_per_capita",
    "CO2 per Capita" = "co2_per_capita",
    "Plastic Waste per Capita" = "plastic_waste_per_capita"))

selectInput("y_variables", "Select Y Variable",
            choices = c(
    "GDP per Capita" = "gdp_per_capita",
    "CO2 per Capita" = "co2_per_capita",
    "Plastic Waste per Capita" = "plastic_waste_per_capita"))

radioButtons("radio_trans", "Select Transformation",
             choices = list("None" = 1,
                            "Squared" = 2,
                            "Log" = 3,
                            "Reciprocal" = 4),
             selected = 1)
```

```{r}
#| label: Interactive GDP & Environmental Indicators Bubble Plot
#| message: false
#| warning: false

plotlyOutput("reac_bubble")
```

<!-- #### 📊 GDP vs Plastic Waste Per Capita (2019 & 2020) -->

<!-- ```{r} -->
<!-- #| warning: false -->
<!-- #| label: fig-rq1-facet -->
<!-- #| fig-cap: "Relationship between GDP per capita and CO2 waste per capita across countries in 2019 and 2020, with separate panels for each year." -->
<!-- #| fig-alt: "Faceted scatterplot showing GDP per capita versus CO2 waste per capita for countries in 2019 and 2020. Each panel includes a fitted trend line. The relationship appears positive in both years, allowing comparison of how the pattern changes between 2019 and 2020." -->

<!-- ggplot( -->
<!--   meta_set %>% -->
<!--     filter(year %in% c(2019, 2020), parent_company == "Grand Total"), -->
<!--   aes(x = gdp_per_capita, y = co2_per_capita) -->
<!-- ) + -->
<!--   geom_point(color = "#2C7FB8", alpha = 0.7, size = 2.5) + -->
<!--   geom_smooth(method = "loess", se = FALSE, color = "#D95F0E", linewidth = 1) + -->
<!--   scale_x_log10(labels = scales::label_dollar()) + -->
<!--   scale_y_log10(labels = scales::label_number()) + -->
<!--   facet_wrap(~year) + -->
<!--   labs( -->
<!--     title = "GDP per Capita vs CO2 Waste per Capita", -->
<!--     subtitle = "Comparison of country-level relationships in 2019 and 2020", -->
<!--     x = "GDP per Capita", -->
<!--     y = "CO2 Waste per Capita" -->
<!--   ) + -->
<!--   theme_ft_rc() -->
<!-- ``` -->

# Inside look at Luxembourg

### Why Luxembourg
A country we found very interesting in this data was Luxembourg. We found it fascinating that the GDP per capita was the largest in our dataset (by a considerable margin) and it was also third in CO2 emissions per capita. Yet despite this the whole country has a population that hardly exceeds 600,000 people.


### Luxembourg Map
```{r}
#| label: Leaflet Map
library(leaflet)
library(scales)

lux <- model_data |> filter(country.x == "Luxembourg")

leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = 7.5, lat = 50.0, zoom = 6) |>
  addCircleMarkers(
    lng = 6.1296, lat = 49.8153,
    radius = 12,
    color = "#e63946", fillColor = "#e63946", fillOpacity = 0.8,
    popup = paste0(
      "<b>Luxembourg (2019)</b><br>",
      "Population: ", comma(lux$population), "<br>",
      "GDP per Capita: $", comma(round(lux$gdp_per_capita)), "<br>",
      "CO₂ per Capita: ", round(lux$co2_per_capita, 2), " tonnes<br>",
      "Plastic Waste per Capita: ", round(lux$plastic_waste_per_capita * 1e6, 2), " kg"
    )
  )

```



### GDP vs CO₂ Per Capita Over Time (1970–2022)

```{r}
#| label: fig-gapminder-animated
#| warning: false

highlight_countries <- c("Luxembourg", "United States", "China", "India",
                         "Germany", "Japan", "United Kingdom", "Brazil")

highlight_colors <- c(
  "Luxembourg"     = "#e63946",
  "United States"  = "#1f77b4",
  "China"          = "#ff7f0e",
  "India"          = "#2ca02c",
  "Germany"        = "#9467bd",
  "Japan"          = "#8c564b",
  "United Kingdom" = "#17becf",
  "Brazil"         = "#bcbd22"
)

bg_data <- gapminder_co2 |> filter(!country %in% highlight_countries)
hl_data <- gapminder_co2 |>
  filter(country %in% highlight_countries) |>
  mutate(lux_label = ifelse(country == "Luxembourg", "Luxembourg", ""))

plot_ly() |>
  add_trace(
    data      = bg_data,
    x         = ~gdp_per_capita,
    y         = ~co2_per_capita,
    size      = ~population,
    frame     = ~year,
    type      = "scatter",
    mode      = "markers",
    name      = "Other countries",
    hoverinfo = "text",
    text      = ~paste0("<b>", country, "</b><br>",
                        "GDP/cap: $", scales::comma(round(gdp_per_capita)), "<br>",
                        "CO₂/cap: ", round(co2_per_capita, 2), " t"),
    marker    = list(color = "rgba(200,200,200,0.25)",
                     sizemode = "diameter",
                     line = list(color = "rgba(150,150,150,0.1)", width = 0.5))
  ) |>
  add_trace(
    data         = hl_data,
    x            = ~gdp_per_capita,
    y            = ~co2_per_capita,
    size         = ~population,
    color        = ~country,
    colors       = highlight_colors,
    frame        = ~year,
    type         = "scatter",
    mode         = "markers+text",
    text         = ~lux_label,
    textposition = "top right",
    hovertemplate = "%{customdata}<extra></extra>",
    customdata   = ~paste0("<b>", country, "</b><br>",
                           "GDP/cap: $", scales::comma(round(gdp_per_capita)), "<br>",
                           "CO₂/cap: ", round(co2_per_capita, 2), " t<br>",
                           "Pop: ", scales::comma(round(population / 1e6, 1)), "M"),
    marker = list(sizemode = "diameter", opacity = 0.85,
                  line = list(color = "white", width = 1.5))
  ) |>
  layout(
    xaxis  = list(
      title      = "GDP per Capita (USD, PPP)",
      type       = "log",
      tickvals   = c(500, 1000, 2000, 5000, 10000, 20000, 50000, 100000),
      ticktext   = c("$500", "$1K", "$2K", "$5K", "$10K", "$20K", "$50K", "$100K"),
      tickfont   = list(size = 11),
      titlefont  = list(size = 13),
      gridcolor  = "rgba(200,200,200,0.5)"
    ),
    yaxis  = list(title = "CO₂ per Capita (tonnes)", type = "log"),
    title  = list(text = "GDP per Capita vs CO₂ per Capita (1970–2022)",
                  font = list(size = 15)),
    legend = list(title = list(text = "Country")),
    paper_bgcolor = "white",
    plot_bgcolor  = "#f5f5f5",
    height        = 650
  ) |>
  animation_opts(frame = 700, easing = "linear", redraw = FALSE) |>
  animation_slider(currentvalue = list(prefix = "Year: "))
```

```{r}
#| context: server
#| message: false
#| warning: false 

output$reac_bubble <- renderPlotly({
  x_var <- input$x_variables
  y_var <- input$y_variables
  
  plot_data <- model_data %>%
  arrange(desc(population)) %>%
  mutate(
    country.x = as.factor(country.x),
    gdp_level = case_when(
      is.na(gdp_billions) ~ "Other",
      gdp_billions < 100 ~ "Low",
      gdp_billions < 500 ~ "Lower-Mid",
      gdp_billions < 1000 ~ "Upper-Mid",
      gdp_billions >= 1000 ~ "High"
    ),
    gdp_level = factor(
      gdp_level,
      levels = c("Low", "Lower-Mid", "Upper-Mid", "High", "Other")
    ),
     hover_text = paste0(
      country.x,
      "<br>", x_var, ": ", round(.data[[x_var]], 2),
      "<br>", y_var, ": ", round(.data[[y_var]], 2),
      "<br>Population: ", format(population, big.mark = ",")
     )
  ) 
  
  if (input$radio_trans == 2) {
    plot_data <- plot_data %>% 
      mutate(
        x_plot = .data[[x_var]]^2,
        y_plot = .data[[y_var]]^2
      )
  } else if (input$radio_trans == 3) {
    plot_data <- plot_data %>% 
      mutate(
        x_plot = log(.data[[x_var]]),
        y_plot = log(.data[[y_var]])
      )
  } else if (input$radio_trans == 4) {
    plot_data <- plot_data %>% 
      mutate(
        x_plot = 1 / .data[[x_var]],
        y_plot = 1 / .data[[y_var]]
      )
  } else {
    plot_data <- plot_data %>% 
      mutate(
        x_plot = .data[[x_var]],
        y_plot = .data[[y_var]]
      )
  }
  
  nice_labels <- c(
    gdp_per_capita = "GDP per Capita",
    co2_per_capita = "CO2 per Capita",
    plastic_waste_per_capita = "Plastic Waste per Capita"
  )
  
  bubble_plot <- ggplot(plot_data, aes(
    x = x_plot,
    y = y_plot,
    size = population,
    color = gdp_level,
    text = hover_text
  )) +
    geom_point(alpha = 0.65) +
    scale_size(range = c(2, 14), name = "Population") +
    scale_color_viridis_d(option = "A", end = 0.9, name = "GDP Level") +
    theme_ipsum_rc() +
    labs(
      x = nice_labels[x_var],
      y = nice_labels[y_var],
      title = "Interactive GDP & Environmental Indicators Bubble Plot",
      subtitle = "2019 countries with population size"
    ) +
    theme(
      plot.title = element_text(size = 14)
    )

  ggplotly(bubble_plot, tooltip = "text")
})
```

